Spatial and temporal variability in economic availability of residual biomass

### 2021-10-27 analysis/manuscript.Rmd
2021-10-27

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     number_sections: yes
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    includes:
      in_header: "preamble.tex"
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '80%'}
knitr::include_graphics("../img/canada.png")
```

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 000 Mg/a  corresponding to a small wood pellet
mill, 400 000 Mg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 000 Mg/a  corresponding to a large ethanol plant.

```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'}
knitr::include_graphics("../img/study_area.png")
```

@@

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| no road           |     2  |       2  |           2   |

@@

Figure 6 illustrates the road network in Alberta and estimates of one-way haul
time based on the average of loaded and unloaded speeds from Table ??. The
road network in the White Area of the province is very dense and largely arranged
as a grid related to the Alberta Township Survey System.  Most of the roads
in the White Area are public.  The road network in the Green Area is much
less dense and many of the roads are built and maintained by resource industries 
(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018.
QGIS zonal statistics were used to calculate the median network travel time to
each township, from each of our three towns.  The one-way travel time in Figure
6 was limited to 6 hours because this equates to a 13 hour cycle time if
you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
vehicles are not permitted to drive more than 13 hours in a day.  **technically
we could take the picture out to 6.5 hours, as the loading and unloading time
does not count as driving time.  There is an on-duty limit of 14 hours**.

@@

```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
knitr::include_graphics("../img/ecozonesAlex.png")
```

@@

## Distributional results
```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}

cyclePlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg )") +
  labs(y = "marginal cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,14))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
mcPlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg )") +
  labs(y = "marginal cost ($·Mg^-1^ )") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,250)) 
```

@@

aveave

Figure ?? displays the marginal cycle time for each of the locations over the
16 years of the study period.  The year 2009 had the median overall harvest 
level in Alberta over that period, and is coloured red in order to illustrate
the marginal cycle time for a median year.

@@

```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
knitr::include_graphics("../img/wcontours.png")
```
# Conclusions

# Data availability

# Acknowledgements

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990-2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---
\raggedright
<!-- \fontfamily{cmr} -->
<!-- \fontsize{12}{22} -->
<!-- \fontseries{b} -->
<!-- \selectfont -->

@@

@@

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200 Tg/a  corresponding to a small wood pellet
mill, 400 Tg/a  corresponding to a large pellet mill or a small ethanol
plant, and 800 Tg/a  corresponding to a large ethanol plant.

@@

@@

Figure 6 illustrates the road network in Alberta and estimates of one-way haul
time based on the average of loaded and unloaded speeds from Table ??. The
road network in the White Area of the province is very dense and largely arranged
as a grid related to the Alberta Township Survey System.  Most of the roads
in the White Area are public.  The road network in the Green Area is much
less dense and many of the roads are built and maintained by resource industries 
(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
generated using the QGIS Network Analysis Toolbox 3 plugin (QNEAT3) created by 
Raffler @Raffler2018,
specifically its iso-area as interpolation (from point) algorithm.  In our case,
we used the algorithm to determine the minimum one-way travel time along the road network 
from each
of our centres to each cell of a provincial level raster at a 200 m resolution.  Travel time
was calculated based on the harmonic mean of loaded and unloaded times for each
road segment.  Off-road travel was assumed to be at 5 km/h.  The underlying
algorithm for travel time minimization is that of Dijkstra @Dijkstra1959.

QGIS zonal statistics were used to calculate the median network travel time to
each township, from each of our three centres.  The one-way travel time in Figure
6 was limited to 6 hours because this equates to a 13 hour cycle time if
you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
vehicles are not permitted to drive more than 13 hours in a day.  **technically
we could take the picture out to 6.5 hours, as the loading and unloading time
does not count as driving time.  There is an on-duty limit of 14 hours**.

@@

@@

## Distributional results
```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by plant location.  Each curve represents one year.  The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}

cyclePlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Tg )") +
  labs(y = "marginal cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
   geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,14))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.  Each curve represents one year.  The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
mcPlot <- 
  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (Tg )") +
  labs(y = "marginal cost ($·Mg^-1^ )") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,250)) 
```

@@

ave <- aveave %>%
  gt() %>%
  tab_header(
    title = "summary stats"
    ) %>%
  fmt_number(
    columns = c(capacity,aveavecost,sdavecost,cv),
    suffixing = TRUE,
    n_sigfig = 3
  ) %>% 
  cols_label(location = "Location",
             capacity = "Capacity (Tg/a)",
             aveavecost = "Mean ($·Mg^-1^)", 
             sdavecost = "Std. Dev. ($·Mg^-1^)", 
             cv = "C.V.")
```

Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
for each of the 16 years in the study period, and for the 3 alternative plant 
locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
years of minimum, median, and maximum provincial harvest volume in the period. 
The inter-year variation in marginal costs increases as the cumulative amount
of residue collected increases.  We can also see that the relative position of
the years in terms of cost varies between the plant locations:  *e.g.* the year 
2005 is not as good for Hinton, as it was for Edson and Whitecourt.

Figure 11 is another way of presenting the year to year variation in the 
distribution of costs between the plant locations and assumed capacity: the
boxplots display the quartiles and the minimum and maximum costs require to
supply a plant of a given capacity at each of the locations in each of the
years.  Here again, we can see the variability increasing as plant capacity
increases.

Figure 12 shows the average cost ($·Mg^-1^) for  each location, capacity, and
year.  Overall, the average costs in Edson and Whitecourt are quite similar.
Hinton is usually more expensive across years and capacities.  The differences
become clearer at a capacity of 800 Tg, where Whitecourt is the low cost location
for most of the years.

Table ?? presents summary statistics over the 16 year for the locations and 
capacities.  It is clear that average costs and the inter-annual variability of 
costs increase with increasing capacity.  It is clear that the variation is costs
for Hinton is greater than for Whitecourt and Edson, particularly for the 800
Tg/a capacities.

Figure 13 shows the areas that would have been needed to supply a plants of 
varying capacities in Whitecourt in the year of median harvest (2009) in 
relation to the proportional township area harvested.  There is some residue 
being collected from the White area, but most of it is being collected from the 
green area.

Figure 14 compares the supply areas for 2005, 2009, and 1990 (high, median, and
low harvest years).  The area required to supply the mills increases by a lot
in low timber harvest years, particularly for the 800 Tg/a capacity.  Similar
maps could be produced for the other locations and all 16 years, but that is
unecessary here.

<!-- A table aligned on decimal point. -->

<!-- \begin{tabular}{S[table-format=3.2]}% syntax for siunitx v2; for v1 use "tabformat" -->
<!-- 555 \\ -->
<!-- 7.77 \\ -->
<!-- 99.9 -->
<!-- \end{tabular} -->

<!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
<!-- 16 years of the study period.  The year 2009 had the median overall harvest  -->
<!-- level in Alberta over that period, and is coloured red in order to illustrate -->
<!-- the marginal cycle time for a median year. -->

# Conclusions

1. Spatial resolution between Yemshanov and FPInnovations

1. Demonstrated spatial and temporal variation in 
  availability of residue supply for plant
  locations of different capacities.  Mention average and marginal costs.

2. Demonstrated use of publicly available data to quantify this stuff.

3. Residue supply to large plants is more expensive and more variable than
 to smaller plants.

Perhaps it makes sense to transport to smaller plants (e.g. pellets), and
then aggregate at a larger plant (e.g. ethanol).  Pellets are cheaper to 
transport than comminuted biomass.

# Data availability {-}

**Put the compiled data on ERA.  Point out links to other information**

# Acknowledgements {-}

\newpage

<div id="refs"></div>

\newpage

# Tables {-}

Table:  Road speeds (km/h).

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| no road           |     5  |       5  |           5   |

Table:  blah blah blah

```{r ave, echo=FALSE}
ave
```

\newpage

# Figures {-}

```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '100%'}
knitr::include_graphics("../img/canada.png")
```

<!-- ```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'} -->
<!-- knitr::include_graphics("../img/study_area.png") -->
<!-- ``` -->

<!-- ```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'} -->
<!-- knitr::include_graphics("../img/ecozonesAlex.png") -->
<!-- ``` -->

@@

\newpage
\blandscape
```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
knitr::include_graphics("../img/wcontours.png")
```
\elandscape
